---
format: gfm
---

# nonbgfa

<!-- badges: start -->
[![R-CMD-check](https://github.com/BIFX547-26/non-b-gfa-johnsonra/actions/workflows/R-CMD-check.yml/badge.svg)](https://github.com/BIFX547-26/non-b-gfa-johnsonra/actions/workflows/R-CMD-check.yml)
<!-- badges: end -->

```{r setup}
#| include: false
library(knitr)
library(GenomicRanges)
```

**nonbgfa** is an R package for finding non-B DNA-forming motifs in genomic
sequences.  It wraps the [GFA suite](https://nonb-abcc.ncifcrf.gov/apps/site/default)
developed at NCI-Frederick / Frederick National Laboratory for Cancer Research,
exposing the C algorithms via a `.Call()` interface so results are returned
directly as R `data.frame` or `GenomicRanges` objects.

Non-B DNA structures deviate from the canonical right-handed Watson–Crick
double helix and are implicated in genomic instability, DNA recombination
hotspots, and elevated mutation rates in cancer.

## Motif types

| Function | Motif | Structural form |
|---|---|---|
| `find_ir()` | Inverted repeats | Cruciform DNA |
| `find_mr()` | Mirror repeats | Triplex (H-DNA) |
| `find_dr()` | Direct repeats | Slipped-strand DNA |
| `find_gq()` | G-quadruplexes | G4 / tetraplex DNA |
| `find_zdna()` | Z-DNA | Left-handed helix |
| `find_str()` | Short tandem repeats | Microsatellites |
| `find_apr()` | A-phased repeats | Bent DNA |

## Installation

```r
# Install from GitHub
# install.packages("remotes")
remotes::install_github("abcsFrederick/non-b-gfa", ref = 'Rpkg')
```

The package requires a C compiler (provided by
[Rtools](https://cran.r-project.org/bin/windows/Rtools/) on Windows, Xcode
command-line tools on macOS, or `build-essential` on Linux).

## Quick start

```{r quick-start-all}
library(nonbgfa)

# Use the bundled example sequence
fasta <- system.file("extdata", "gfa_test.fasta", package = "nonbgfa")

# Run all seven finders at once
results <- find_nonb(fasta)
sapply(results, nrow)
```

```r
# Or call individual finders
ir  <- find_ir(fasta)
gq  <- find_gq(fasta)
head(ir)
```

```{r quick-start-singles}
#| echo: false
ir  <- find_ir(fasta)
head(ir) |>
    kable()
```


## Output columns

Every finder returns a `data.frame` with these columns:

| Column | Description |
|---|---|
| `seq_name` | Sequence identifier from the FASTA `>` header |
| `start` | Start position (1-based, inclusive) |
| `end` | End position (1-based, inclusive) |
| `strand` | `"+"` (sense) or `"-"` (antisense) |
| `length` | Repeat-unit length (bp); G-run size for GQ |
| `spacer` | Spacer between repeat halves; KV score for Z-DNA |
| `num_repeats` | Number of times the unit repeats; permutations for MR |
| `remainder` | Partial repeat (DR); min-loop boundary (IR/MR); island count (GQ) |
| `subset` | `TRUE` if locus qualifies as cruciform/triplex/slipped/high-KV |

Pass `format = "GRanges"` to any finder for a
[`GenomicRanges::GRanges`](https://bioconductor.org/packages/GenomicRanges) object
(requires the **GenomicRanges** Bioconductor package).

```{r}
find_ir(fasta, format = 'GRanges') |>
    head()
```

## Parameter defaults

All parameters match the original `gfa` CLI defaults and the Non-B DB website.

| Motif | Parameter | Default | Meaning |
|---|---|---|---|
| IR | `minIRrep` | 6 | Min arm length (bp) |
| IR | `maxIRspacer` | 100 | Max spacer (bp) |
| IR | `shortIRcut` | 9 | Arms ≤ this length require tight spacer |
| IR | `shortIRspacer` | 4 | Spacer limit for short IRs |
| MR | `minMRrep` | 10 | Min half-length (bp) |
| MR | `maxMRspacer` | 100 | Max spacer (bp) |
| DR | `minDRrep` / `maxDRrep` | 10 / 300 | Repeat unit size range (bp) |
| DR | `maxDRspacer` | 10 | Max spacer (bp) |
| GQ | `minGQrep` | 3 | Min consecutive G's per run |
| GQ | `maxGQspacer` | 7 | Max spacer between G-runs (bp) |
| Z-DNA | `minZlen` | 10 | Min alternating pur/pyr run (bp) |
| STR | `minSTR` / `maxSTR` | 1 / 9 | Repeat unit size range (bp) |
| STR | `minSTRbp` | 10 | Min total locus length (bp) |
| APR | `minAPRlen` / `maxAPRlen` | 3 / 9 | A-tract length range (bp) |
| APR | `minATracts` | 3 | Min A-tracts per repeat |

See `?find_nonb` and individual function help pages for the full parameter list.

## Multi-sequence input

Functions will accept either a single named character vector or a multi-sequence FASTA file:

```r
# input from fasta file
fasta_ir <- read_fasta(fasta) |>
    find_ir()

# Inline sequences
inline <- c(my_gene = "atcgatcgatcgatcgatcg") |>
    find_ir()
```

## Citation

Please cite the original GFA tool:

> Cer RZ, Donohue DE, Mudunuri US, Temiz NA, Loss MA, Starner NJ,
> Halusa GN, Volfovsky N, Yi M, Luke BT, Bacolla A, Collins JR,
> Stephens RM. (2013) Non-B DB v2.0: a database of predicted non-B
> DNA-forming motifs and its associated tools. *Nucleic Acids Research*,
> 41(D1):D94–D100. <https://doi.org/10.1093/nar/gks955>

## Original C tool

The original `gfa` command-line tool source and Makefile are preserved in
`inst/legacy/` for reference.  To build the standalone binary:

```sh
cd inst/legacy
make
./gfa -skipWGET -seq ../../inst/extdata/gfa_test.fasta -out gfa_test
```
