---
title: "Introduction to nonbgfa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to nonbgfa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

**nonbgfa** identifies sequences associated with non-B DNA-forming motifs in
genomic FASTA sequences.  It wraps the GFA suite originally developed at
NCI-Frederick / Frederick National Laboratory for Cancer Research.

Non-B DNA structures deviate from the canonical right-handed Watson–Crick
double helix and are associated with genomic instability, recombination
hotspots, and elevated mutation rates in cancer.  Seven motif classes are
supported:

| Function | Motif type | Structural form |
|---|---|---|
| `find_ir()` | Inverted repeats | Cruciform DNA |
| `find_mr()` | Mirror repeats | Triplex (H-DNA) |
| `find_dr()` | Direct repeats | Slipped-strand DNA |
| `find_gq()` | G-quadruplexes | G4 / tetraplex DNA |
| `find_zdna()` | Z-DNA | Left-handed helix |
| `find_str()` | Short tandem repeats | Microsatellites |
| `find_apr()` | A-phased repeats | Bent DNA |

## Installation

```r
# Install from GitHub (package not yet on CRAN)
# install.packages("pak")
pak::pak("johnsonra/non-b-gfa")
```

## Quick start

The package ships with a small test sequence that exercises all seven finders.

```{r load}
library(nonbgfa)
fasta <- system.file("extdata", "gfa_test.fasta", package = "nonbgfa")
```

### Run all finders at once

`find_nonb()` runs every enabled finder and returns a named list of
`data.frame`s, one per motif type.

```{r find_nonb}
results <- find_nonb(fasta)
# Number of loci found per motif type
sapply(results, nrow)
```

### Individual finders

Each finder can also be called directly:

```{r individual}
ir  <- find_ir(fasta)
gq  <- find_gq(fasta)
str_res <- find_str(fasta)

head(ir[, c("seq_name", "start", "end", "length", "spacer", "subset")])
```

## Output columns

Every finder returns a `data.frame` with these columns:

| Column | Description |
|---|---|
| `seq_name` | Sequence identifier from the FASTA `>` header |
| `start` | Start position (1-based, inclusive) |
| `end` | End position (1-based, inclusive) |
| `strand` | `"+"` (sense) or `"-"` (antisense) |
| `length` | Repeat-unit length (bp); G-run size for G4 |
| `spacer` | Spacer between repeat halves; KV score for Z-DNA |
| `num_repeats` | Times the unit repeats; permutation count for MR |
| `remainder` | Partial repeat at motif end (DR); min-loop (IR/MR); island count (GQ) |
| `subset` | `TRUE` if locus qualifies as cruciform / triplex / slipped / high-KV |

## Adjusting parameters

All parameters correspond directly to the original `gfa` CLI switches and
default to the same values used by the Non-B DB website.

```{r params}
# Stricter G-quadruplex search: >=4 G-runs, spacer <=3 bp
strict_gq <- find_gq(fasta, minGQrep = 4, maxGQspacer = 3)
nrow(strict_gq)

# Longer inverted-repeat arms
long_ir <- find_ir(fasta, minIRrep = 12, maxIRspacer = 50)
nrow(long_ir)
```

### Key parameter defaults

| Motif | Parameter | Default | Meaning |
|---|---|---|---|
| IR | `minIRrep` | 6 | Min arm length (bp) |
| IR | `maxIRspacer` | 100 | Max spacer (bp) |
| IR | `shortIRcut` | 9 | Arms <= this length require tight spacer |
| IR | `shortIRspacer` | 4 | Spacer limit for short IRs |
| IR | `minCruciformRep` | 6 | Min arm for cruciform flag |
| IR | `maxCruciformSpacer` | 4 | Max spacer for cruciform flag |
| MR | `minMRrep` | 10 | Min half-length (bp) |
| MR | `maxMRspacer` | 100 | Max spacer (bp) |
| MR | `minTriplexYRpercent` | 10 | Min Y/R% for triplex flag |
| MR | `maxTriplexSpacer` | 8 | Max spacer for triplex flag |
| DR | `minDRrep` | 10 | Min unit length (bp) |
| DR | `maxDRrep` | 300 | Max unit length (bp) |
| DR | `maxDRspacer` | 10 | Max spacer (bp) |
| DR | `maxSlippedSpacer` | 0 | Max spacer for slipped flag |
| GQ | `minGQrep` | 3 | Min consecutive G's per run |
| GQ | `maxGQspacer` | 7 | Max spacer between G-runs (bp) |
| Z-DNA | `minZlen` | 10 | Min alternating pur/pyr run (bp) |
| STR | `minSTR` | 1 | Min repeat unit (bp) |
| STR | `maxSTR` | 9 | Max repeat unit (bp) |
| STR | `minSTRbp` | 10 | Min total locus length (bp) |
| APR | `minAPRlen` | 3 | Min A-tract length (bp) |
| APR | `maxAPRlen` | 9 | Max A-tract length (bp) |
| APR | `minATracts` | 3 | Min A-tracts per repeat |

## GenomicRanges output

Any finder accepts `format = "GRanges"` to return a
[`GenomicRanges::GRanges`](https://bioconductor.org/packages/GenomicRanges)
object.  This requires the **GenomicRanges** Bioconductor package.

```r
# BiocManager::install("GenomicRanges")  # if needed
ir_gr <- find_ir(fasta, format = "GRanges")
ir_gr
```

## Skipping motif types

`find_nonb()` accepts a `skip` argument to omit one or more finders:

```{r skip}
# Run everything except STR and APR
fast_results <- find_nonb(fasta, skip = c("STR", "APR"))
names(fast_results)
```

## Working with multiple sequences

Pass a multi-sequence FASTA or a named character vector directly:

```{r multi}
seqs <- read_fasta(fasta)
cat("Sequence length:", nchar(seqs), "bp\n")
```

## Citation

Please cite the original GFA tool when using this package:

> Cer RZ, Donohue DE, Mudunuri US, Temiz NA, Loss MA, Starner NJ,
> Halusa GN, Volfovsky N, Yi M, Luke BT, Bacolla A, Collins JR,
> Stephens RM. (2013) Non-B DB v2.0: a database of predicted non-B
> DNA-forming motifs and its associated tools. *Nucleic Acids Research*,
> 41(D1):D94–D100. <https://doi.org/10.1093/nar/gks955>
